# Storms in the United States 1950 - 2011
========================================================================================================

As stated in the introduction to the course assignment, storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

In this report we try to answer some basic questions about severe weather events, specifically which weather events cause the most human and economic damage. To do that we use the dataset created by the U.S. National Oceanic and Atmospheric Administration's (NOAA) that contains characteristics of major storms and weather events in the United States, including estimates of any fatalities and injuries, and property and crop damage. Our analysis has revealed that tornadoes, heat and floods are the main causes for fatalities, tornadoes are causing the most injuries, and floods, tornadoes, thunderstorms and hails are causing the most economic damage in the US.

## Data Processing

The data has been processed and analysed using the following environment settings:

```{r session_info}
sessionInfo()
```

Following libraries that are used for the analysis, and need installing and loading:

```{r install_packages, results='hide', message=F, warning=F}
## install dplyr if not already done so
if (("dplyr" %in% (installed.packages())) == F) {
        install.packages("dplyr")
}
## install quantmod if not already done so
if (("quantmod" %in% (installed.packages())) == F) {
        install.packages("quantmod")
}
## install quantmod if not already done so
if (("ggplot2" %in% (installed.packages())) == F) {
        install.packages("ggplot2")
}

library(dplyr)
library(quantmod)
library(ggplot2)
```

We download the data for the analysis from the [course file](https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2) repository as indicated in the assessment description:

```{r read_data, cache=TRUE, results='hide', message=F, warning=F}
## the name of the dataset file
data_file <- "StormData.csv"

## download dataset file if not already done so
if (!file.exists(data_file)) {
        ## download the dataset
        download.file(url = "https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", 
                      destfile = "StormData.csv.bz2", method = "curl")
        system("bunzip2 StormData.csv.bz2")
}

## read the entire dataset into a data frame
rawdata <- read.csv(file = data_file)
```

Now we will take a look at the data and make some transformations.
First, check the structure of the data frame:

```{r data_str}
str(rawdata)
```

For our analysis we're interested in the following columns as they are described in the code book (National Climatic Data Center's [record layout document](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc) as referenced on the [Investigative Reporers & Editors](http://ire.org/nicar/database-library/databases/storm-events/) website.):

| Column        | Description |
| ------------- |:-----------:|
| BGN_DATE | The date when the event began |
| EVTYPE | The event type |
| FATALITIES | The number of direct fatalities |
| INJURIES | The number of direct injuries |
| PROPDMG | The damage to the properties |
| PROPDMGEXP | The exponent used to calculate the total amount of property damage |
| CROPDMG | The damage to the crops |
| CROPDMGEXP | The exponent used to calculate the total amount of crop damage |

Using the following code we select the data we're interested in, and transform values to formats that make the analysis easier:

```{r select_convert_data}
## create function that will convert character vectors to numeric exponents
convert_exp <- function(exponent) {
        sapply(exponent, function(exponent) {
                ## try converting to numeric
                n_exp = as.numeric(exponent)
                ## if the result is NA...
                if (is.na(n_exp)) {
                        ## ...convert Factor to character vector, and
                        s_exp = toupper(as.character(exponent))
                        ## ...convert according to this table:
                        ## Hundred (H), Thousand (K), Million (M), Billion (B)
                        n_exp = ifelse(s_exp == "H", 2,
                                ifelse(s_exp == "K", 3,
                                ifelse(s_exp == "M", 6,
                                ifelse(s_exp == "B", 9, 0))))
                }
                return(n_exp)
        })
}
## using the function convert_exp convert the columns PROPDMGEXP and CROPDMGEXP into numeric exponents
data <- tbl_df(rawdata) %>%
        ## select the data we're interested in
        select(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
        ## transform the data into formats we need for our analysis
        mutate(YEAR = as.numeric(format(as.Date(as.character(BGN_DATE), format="%m/%d/%Y %H:%M:%S"), "%Y")),
               EVTYPE = as.character(EVTYPE), 
               ## convert exponents using function convert_exp
               PROPDMGEXP = convert_exp(PROPDMGEXP), 
               CROPDMGEXP = convert_exp(CROPDMGEXP),
               ## calculate damage using the converted numeric exponents
               PROPDMGCASH = PROPDMG * 10^PROPDMGEXP,
               CROPDMGCASH = CROPDMG * 10^CROPDMGEXP,
               ## calculate total economic damage
               TOTALDMGCASH = PROPDMGCASH + CROPDMGCASH,
               ## drop BGN_DATE column
               BGN_DATE = NULL)
```

Check the values for some of the columns by printing out the unique values:

```{r unique_values}
unique(data$EVTYPE)
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

The values for PROPDMGEXP and CROPDMGEXP seem reasonable, but there are some values in the column EVTYPE that don't make any sense:

```{r useless_evtype}
data$EVTYPE[grepl("Summary", data$EVTYPE)]
```

The values are all summaries of some other dates. Checking the values of the raw data:

```{r check_rawdata}
summaries <- rawdata[grepl("Summary", rawdata$EVTYPE), ]

## Check whether this data contains any useful data:
sum(summaries$FATALITIES)
sum(summaries$INJURIES)
sum(summaries$PROPDMG)
sum(summaries$CROPDMG)
```

All these columns contain zero (0) values, so we can discard them:

```{r discard_useless}
data <- filter(data, grepl("Summary", EVTYPE) == FALSE)
```

Considering that the data starts from years 1950 - 2011 we should adjust the damage sums for inflation:

```{r adjusting_inflation}
invisible(getSymbols("CPIAUCSL", src='FRED'))
avg_cpi <- apply.yearly(CPIAUCSL, mean)
conversion_factors <- as.data.frame(avg_cpi / as.numeric(avg_cpi['2011']))
conversion_factors$YEAR = as.numeric(format(as.Date(rownames(conversion_factors)), '%Y'))
colnames(conversion_factors)[1] = 'FACTOR'

data = merge(data, conversion_factors, by='YEAR')
data$TOTALDMGCASH = data$TOTALDMGCASH * data$FACTOR
```

Beside the summary rows, we've seen values that are slightly different, but have very similar meaning in terms of the type of the event, e.g. "FLOOD"" and "FLOODING", or contain typos, e.g. "THUNDER" and "THUNER"" . We will try to merge them into a smaller set of event types to make the analysis exacter:

```{r merge_evtypes, cache=TRUE}
## create function that will merge similar event type names
merge_evtypes <- function(evtype) {
        UPR = toupper(gsub("^\\s+|\\s+$", "", evtype))
        new_evtype = ifelse(grepl("HURRICANE|TYPHOON", UPR), "HURRICANE",
                ifelse(grepl("(THUNDER|THUNER|THUDER|THUNDEER|TUNDER|TSTM).?", UPR) & !grepl("NON", UPR), "THUNDERSTORM", 
                ifelse(grepl("(RECORD|HEAVY|EXCESSIVE|UNSEASONAL).?RAIN.*", UPR) & !grepl("THUNDERSTORM|TSTM|NON", UPR), "HEAVY RAIN",
                ifelse(grepl("STORM SURGE", UPR), "STORM SURGE", 
                ifelse(grepl("(GRASS|WILD|FOREST).?FIRE", UPR), "WILD FIRE",
                ifelse(grepl("HEAT", UPR), "HEAT", 
                ifelse(grepl("COLD|FREEZE|FROST", UPR) & !grepl("AIR|FOG|WET|SNOW", UPR), "", 
                ifelse(grepl("(HIGH|BLOWING|STRONG).?WIND", UPR), "",
                ifelse(grepl("FLOOD", UPR), "FLOOD",
                ifelse(grepl("TROPICAL.?STORM", UPR), "TROPICAL STORM", 
                ifelse(grepl("FOG", UPR), "FOG",
                ifelse(grepl("WINTRY|WINTER", UPR), "WINTER WEATHER",
                ifelse(grepl("HAIL.?", UPR), "HAIL",
                ifelse(grepl("BLOWING.?SNOW", UPR), "BLOWING SNOW",
                ifelse(grepl("DRY.?MICROBURST", UPR), "DRY MICROBURST",
                ifelse(grepl("FREEZING.?(DRIZZLE|RAIN|SPRAY)", UPR), "FREEZING RAIN",
                ifelse(grepl("(RECORD|HEAVY).?SNOW", UPR), "HEAVY SNOW",
                ifelse(grepl("ICE|ICY", UPR), "ICE",
                ifelse(grepl("LIGHTNING", UPR), "LIGHTNING",
                ifelse(grepl("RIP.?CURRENT.?", UPR), "RIP CURRENT",
                ifelse(grepl("TORNADO|TORNDAO", UPR), "TORNADO",
                ifelse(grepl("GUSTY.*WIND.*", UPR), "GUSTY WIND",
                ifelse(grepl("COLD.*WET.*", UPR), "COLD AND WET",
                ifelse(grepl("EXTREME.*WIND.*(DAMAGE|STORM)*", UPR), "EXTREME WIND",
                ifelse(grepl("HIGH.*SURF.*", UPR), "HIGH SURF",
                ifelse(grepl("LAKE.*EFFECT.*SNOW", UPR), "LAKE EFFECT SNOW",
                ifelse(grepl("LAND.*(SLIDE|SLIDES|SLUMP|SPOUT)", UPR), "LAND SLIDE",
                ifelse(grepl("LIGHTING|LIGHTNING|LIGNTNING", UPR), "LIGHTNING",
                ifelse(grepl("MICROBURST", UPR), "MICROBURST",
                ifelse(grepl("MUD.*SLIDE.*", UPR), "MUD SLIDE",
                ifelse(grepl("WATERSPOUT.*", UPR), "WATERSPOUT", UPR)))))))))))))))))))))))))))))))

        return(new_evtype)
}
## merge similar event types
data$EVTYPE <- sapply(data$EVTYPE, merge_evtypes)
```

## Results

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  

Calculate total fatalities and injuries per event type and select the top ten harmful events:

```{r population_damage}
fatalities <- data
        ## group data by event type
        group_by(EVTYPE) %>%
        ## calculate total fatalities per event type
        summarize(FATALITIES = sum(FATALITIES)) %>%
        ## discard data where event type value is empty, or fatalities value is zero
        filter(EVTYPE != "" & FATALITIES > 0) %>%
        ## sort by fatalities in descending order
        arrange(desc(FATALITIES)) %>%
        ## selet the first 10 rows
        slice(1:10)

injuries <- data %>%
        ## group data by event type
        group_by(EVTYPE) %>%
        ## calculate total fatalities per event type
        summarize(INJURIES = sum(INJURIES)) %>%
        ## discard data where event type value is empty, or fatalities value is zero
        filter(EVTYPE != "" & INJURIES > 0) %>%
        ## sort by fatalities in descending order
        arrange(desc(INJURIES)) %>%
        ## selet the first 10 rows
        slice(1:10)
        
```

Plot fatalities by event type:

```{r plot_fatalities}
ggplot(data = fatalities, aes(x = EVTYPE, y = FATALITIES)) +
        geom_bar(position="dodge", stat='identity', colour="black", fill="blue") +
        coord_flip() +
        ggtitle("Total fatalities per weather event type") +
        scale_x_discrete(limits = fatalities$EVTYPE)
```

The plot above shows that the most harmful weather events to human population are tornadoes, heat waves and floods

Plot injuries by event type:

```{r plot_injuries}
ggplot(data = injuries, aes(x = EVTYPE, y = INJURIES)) +
        geom_bar(position="dodge", stat='identity', colour="black", fill="blue") +
        coord_flip() +
        ggtitle("Total injuries per weather event type") +
        scale_x_discrete(limits = injuries$EVTYPE)
```

The plot above shows that the weather event that causes the most injuries are tornadoes

2. Across the United States, which types of events have the greatest economic consequences?  
 
```{r economic_damage}
economic_damage <- data %>%
        group_by(EVTYPE) %>%
        summarize(DAMAGE = sum(TOTALDMGCASH)) %>%
        filter(EVTYPE != "" & DAMAGE > 0) %>%
        arrange(desc(DAMAGE)) %>%
        slice(1:10)
```

Plot economic damage by event type:

```{r plot_injuries}
ggplot(data = economic_damage, aes(x = EVTYPE, y = DAMAGE)) +
        geom_bar(position="dodge", stat='identity', colour="black", fill="blue") +
        coord_flip() +
        ggtitle("Total economic damage per weather event type") +
        scale_x_discrete(limits = economic_damage$EVTYPE)
```

In this plot we can see that the most economically damaging events are floods, tornadoes and thunderstorms.





