# Storms in the United States 1950 - 2011
========================================================================================================

TODO: Synopsis describing the analysis

## Data Processing

The data has been processed and analysed using the following environment settings:

```{r session_info}
sessionInfo()
```

Following libraries that are going to be used for the analysis and need installing and attaching:

```{r install_packages, results='hide', message=F, warning=F}
## install dplyr if not already done so
if (("dplyr" %in% (installed.packages())) == F) {
        install.packages("dplyr")
}

library(dplyr)
```

Download and read the data for the analysis:

```{r read_data, cache=TRUE, results='hide', message=F, warning=F}
## the name of the dataset file
data_file <- "StormData.csv"

## download dataset file if not already done so
if (!file.exists(data_file)) {
        ## download the dataset
        download.file(url = "https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", 
                      destfile = "StormData.csv.bz2", method = "curl")
        system("bunzip2 StormData.csv.bz2")
}

## read the entire dataset into a data frame
rawdata <- read.csv(file = data_file)
```

Now we will take a look at the data and make some transformations.
First, view the structure of the data frame:

```{r data_str}
str(rawdata)
```

For our analysis we're interested in the following columns as described in the [Code Book](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc):

- EVTYPE - The event type
- FATALITIES - The number of direct fatalities
- INJURIES - The number of direct injuries
- PROPDMG - The damage to the properties
- PROPDMGEXP - The exponent used to calculate the total amount of property damage
- CROPDMG - The damage to the crops
- CROPDMGEXP - The exponent used to calculate the total amount of crop damage

Select only data we're interested in, and transform values in formats that make the analysis easier:

```{r select_convert_data}
convert_exp <- function(exponent) {
        sapply(exponent, function(exponent) {
                ## try converting to numeric
                n_exp = as.numeric(exponent)
                ## if the result is NA...
                if (is.na(n_exp)) {
                        ## ...convert Factor to character vector, and
                        s_exp = toupper(as.character(exponent))
                        ## ...convert according to this table:
                        ## Hundred (H), Thousand (K), Million (M), Billion (B)
                        n_exp = ifelse(s_exp == "H", 2,
                                ifelse(s_exp == "K", 3,
                                ifelse(s_exp == "M", 6,
                                ifelse(s_exp == "B", 9, 0))))
                }
                return(n_exp)
        })
}

data <- tbl_df(rawdata) %>%
        select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
        mutate(EVTYPE = as.character(EVTYPE), 
               PROPDMGEXP = convert_exp(PROPDMGEXP), 
               CROPDMGEXP = convert_exp(CROPDMGEXP),
               PROPDMGCASH = PROPDMG * 10^PROPDMGEXP,
               CROPDMGCASH = CROPDMG * 10^CROPDMGEXP,
               TOTALDMGCASH = PROPDMGCASH + CROPDMGCASH)

## print out the first 10 lines to see the result of the transformation above:
head(data, 10)
```

Check the values for some of the columns by printing out the unique values:

```{r unique_values}
unique(data$EVTYPE)
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

The values for PROPDMGEXP and CROPDMGEXP seem reasonable, but there are some values in the column EVTYPE that 
don't make any sense:

```{r useless_evtype}
data$EVTYPE[grepl("Summary", data$EVTYPE)]
```

The values are all summaries of some other dates. Checking the values of the raw data:

```{r check_rawdata}
summaries <- rawdata[grepl("Summary", rawdata$EVTYPE), ]

## Check whether this data contains any useful data:
sum(summaries$FATALITIES)
sum(summaries$INJURIES)
sum(summaries$PROPDMG)
sum(summaries$CROPDMG)
```

All these columns contain zero (0) values, so we can discard them:

```{r discard_useless}
data <- filter(data, grepl("Summary", EVTYPE) == FALSE)
```

Beside the summary rows, we've seen values that are slightly different, but have very similar meaning in terms of the type of the event, e.g. "coastal flood"" and "coastal flooding". We will try to merge them into a smaller set of event types to make the analysis exacter:

```{r merge_evtypes, cache=TRUE}
merge_evtypes <- function(evtype) {
        upr = toupper(evtype)
        new_evtype = ifelse(grepl("HURRICANE|TYPHOON", upr), "HURRICANE",
                ifelse(grepl("(THUNDERSTORM|TSTM).?WIND", upr) & !grepl("NON", upr), "THUNDERSTORM WIND", 
                ifelse(grepl("(HEAVY|EXCESSIVE|UNSEASONAL).?RAIN", upr) & !grepl("THUNDERSTORM|TSTM|NON", upr), "HEAVY RAIN",
                ifelse(grepl("STORM SURGE", upr), "STORM SURGE", 
                ifelse(grepl("(GRASS|WILD|FOREST).?FIRE", upr), "WILD FIRE",
                ifelse(grepl("HEAT", upr), "HEAT", 
                ifelse(grepl("COLD|FREEZE|FROST", upr) & !grepl("AIR|FOG|WET|SNOW", upr), "", 
                ifelse(grepl("(HIGH|BLOWING|STRONG).?WIND", upr), "",
                ifelse(grepl("FLOOD", upr), "FLOOD",
                ifelse(grepl("TROPICAL.?STORM", upr), "TROPICAL STORM", 
                ifelse(grepl("FOG", upr), "FOG",
                ifelse(grepl("WINTRY|WINTER", upr), "WINTER WEATHER", upr))))))))))))

        return(new_evtype)
}
data$EVTYPE <- sapply(data$EVTYPE, merge_evtypes)
```


## Results

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?  
 








