# Storms in the United States Between 1950 and 2011
========================================================================================================

TODO: Synopsis describing the analysis

## Data Processing

The data has been processed and analysed using the following environment settings:

```{r}
sessionInfo()
```

Install and import libraries that are going to be used for the analysis:

```{r install_packages, results='hide', message=F, warning=F}
## install dplyr if not already done so
if (("dplyr" %in% (installed.packages())) == F) {
        install.packages("dplyr")
}

library(dplyr)
```

Download and read the data for the analysis:

```{r read_data, cache=TRUE, results='hide', message=F, warning=F}
## the name of the dataset file
data_file <- "StormData.csv"

## download dataset file if not already done so
if (!file.exists(data_file)) {
        ## check OS
        is_windows <- .Platform$OS.type == "windows"
        ## download the dataset
        download.file(url = "https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2", 
                      destfile = "StormData.csv.bz2",
                      method = ifelse(is_windows, "auto", "curl"))
        system("bunzip2 StormData.csv.bz2")
}

## read the entire dataset into a data frame wth columns of types numeric and character. We will convert the data types
## of some of them according to our needs
rawdata <- read.csv(file = data_file, stringsAsFactors = F)
```

Now we will take a look at the data and make some transformations.
First, the structure of the data frame:
```{r}
str(rawdata)
```

For our analysis we're interested in the following columns:

- EVTYPE - The event type
- FATALITIES - The number of fatalities
- INJURIES - The number of injuries
- PROPDMG - The damage to the properties
- PROPDMGEXP - 
- CROPDMG - The damage to the crops
- CROPDMGEXP - 

Select only data we're interested in:
```{r}
data <- select(tbl_df(rawdata), EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```



## Results











